Exploring SSA Optimization Techniques

### 2014-11-11 doc/paper.tex
Added block inline section

\documentclass[12pt, a4paper]{article}
\usepackage{amsmath, listingsutf8, mathtools}
\lstset{numbers=left}
\title{Exploring SSA Optimization Techniques}
\author{A. D. Blom}
\date{}
\begin{document}
  \maketitle
  \begin{abstract}
In this era of global digitalization, the need for performant software is larger
than ever. For the production of fast software not only 
fast hardware and solidly written code are needed, but also a well-optimizing 
compiler. This paper explores some common optimization techniques based on an 
SSA intermediate language.
  \end{abstract}

@@

  \section{Intermediate representation used}
The intermediate representation (IR) used here is an SSA implentation, 
partially based on the LLVM IR. There are a few differences: the IR used here 
uses a typesystem analogous to C's, and uses a C-like syntax for textual 
representation as well.

@@

Can't use a $\phi$ node system, since it needs \verb+i+ to actually exist in 
memory. It is therefore hard for a front-end to decide which system to use, and 
many default to using the first system all of the time, relying on the 
middle-end to optimize it into a $\phi$ node system. If an \verb+alloca+ 
variable can convert its \verb+load+/\verb+store+ system it shall be considered 
\textit{phiable}.

\subsection{Implementation}
Given a simple, one-block SSA graph:

@@

Can \verb+t1+ be considered phiable? It's been established that an alloca system 
is not phiable if the memory is actually required to exist. This means (naively) 
that a system is not phiable if the \verb+alloca+ instruction is used outside 
its own \verb+load+/\verb+store+ instructions: that is if it is ever used in an 
instruction, except as the first operand of a \verb+store+ or \verb+load+.

@@

For the load in line 7 for example, finding the last store is non-trivial, it 
has in fact got multiple last store instructions, one in \verb+L1+ and one in 
\verb+L2+. It is now actually required to implement a  $\phi$ node. It selects 
the value from L1 if that was its predecessor, and the store from \verb+L2+ if 
that was its predecessor using a $\phi$ node:

@@

\subsection{Implementation}
In order to perform any useful consant folding, the compiler needs to fill in 
constants for variables where possible, so code of the form:

@@

\subsection{Considerations}
\subsubsection{Platform incompatibilities}
There is a way compiler-based constant folding might stand in the way of the 
programmer. Mostly the compiler can do this when folding away instructions 
operating on floating point operands, because different targets may compute 
floating point operations differently. Therefore cross-compilation becomes an 
issue; if a floating point instruction for target $Y$ normally yielding $V_y$, it 
yields $V_x$ when folded away by target $X$, causing different semantics before and 
after optimisation.

@@

\subsection{Implementation}
The implementation of this optimization simply needs to check whether the first
parameter of a \verb+split+ instruction is constant, and convert it into a
\verb+jmp+ accordingly. It also needs to check for the presence of $\phi$ nodes
in the block not covered by the \verb+jmp+ instruction, and remove them accordingly:

@@

>>>>

\documentclass[12pt, a4paper]{article}
\usepackage[T1]{fontenc}
\usepackage{amsmath, listingsutf8, mathtools}

\usepackage{hyperref}
\usepackage{breakurl}

%\hypersetup{}
%\lstset{numbers=left}
\title{Exploring SSA Optimization Techniques}
\author{A. D. Blom}
\date{}

\begin{document}
  \maketitle
  \begin{abstract}
In this era of global digitalization, the need for performant software is larger
than ever. For the production of fast software not only 
fast hardware and solidly written code are needed, but also a well-optimizing 
compiler. This paper explores some common optimization techniques based on an 
SSA intermediate language.
  \end{abstract}

@@

  \section{Intermediate representation used}
The intermediate representation (IR) used here is an SSA implentation, 
partially based on the LLVM IR as described by~LLVM~\cite{llvm_ir}.
There are a few differences: the IR used here  uses a typesystem analogous to
C's, and uses a C-like syntax for textual 
representation as well.

@@

Can't use a $\phi$ node system, since it needs \verb+i+ to actually exist in 
memory. It is therefore hard for a front-end to decide which system to use, and 
many\footnote{At least clang does so: \textit{echo "void foo() \{ int a; \}" |
 clang -x c - -S -emit-llvm -o /dev/stdout}}
default to using the first system all of the time, relying on the 
middle-end to optimize it into a $\phi$ node system. If an \verb+alloca+ 
variable can convert its \verb+load+/\verb+store+ system it shall be considered 
\textit{phiable}.

\subsection[Implementation] {Implementation\footnote{This is the o\_phiable() optimization pass in opt.c in acc} }
Given a simple, one-block SSA graph:

@@

Can \verb+t1+ be considered phiable? It's been established that an \verb+alloca+ system 
is not phiable if the memory is actually required to exist. This means (naively) 
that a system is not phiable if the \verb+alloca+ instruction is used outside 
its own \verb+load+/\verb+store+ instructions: that is if it is ever used in an 
instruction, except as the first operand of a \verb+store+ or \verb+load+.

@@

For the load in line 7 for example, finding the last store is non-trivial, it 
has in fact got multiple last \verb+store+ instructions, one in \verb+L2+ and one in 
\verb+L3+. It is now actually required to implement a  $\phi$ node. It selects 
the value from \verb+L2+ if that was its predecessor, and the store from \verb+L3+ if 
that was its predecessor using a $\phi$ node:

@@

\subsection[Implementation] {Implementation\footnote{This is the o\_cfld() optimization pass in opt.c in acc} }
In order to perform any useful consant folding, the compiler needs to fill in 
constants for variables where possible, so code of the form:

@@

\subsection{Considerations}
\subsubsection{Platform incompatibilities}
There is a way compiler-based constant folding might stand in the way of the 
programmer, as described by~Wikipedia~\cite{cfld_cross}. Mostly the compiler can do this when folding
away instructions  operating on floating point operands, because different targets may compute 
floating point operations differently. Therefore cross-compilation becomes an 
issue; if a floating point instruction for target $Y$ normally yielding $V_y$, it 
yields $V_x$ when folded away by target $X$, causing different semantics before and 
after optimisation.

@@

\subsection[Implementation] {Implementation\footnote{This is the o\_uncsplit() pass in opt.c acc} }
The implementation of this optimization simply needs to check whether the first
parameter of a \verb+split+ instruction is constant, and convert it into a
\verb+jmp+ accordingly. It also needs to check for the presence of $\phi$ nodes
in the block not covered by the \verb+jmp+ instruction, and remove them accordingly:

@@

\section{Block inlining}
\subsection{Problem and implementation}
When a block has only one predecessor and its single predecessor
also has one successor, its instructions can be inlined into the block it succeeds:

\begin{lstlisting}
L0:	...
	jmp(L1);
L1:	int t2 = add((int)0, (int)1);
	jmp(L3);
L3:	...
\end{lstlisting}

Becomes (assuming L1 has no other predecessors):

\begin{lstlisting}
L0:	...
	int t1 = add((int)0, (int)1);
	jmp(L2);
L2:	...
\end{lstlisting}

That way the amount of jumps and blocks is reduced without duplicating instructions.

It's a very trivial optimization but occurs quite a lot, especially considering
the front-end may generate redundant blocks all the time. Consider an infinite
\verb+for+ loop:

\begin{lstlisting}
	for (; cond;)
		;
\end{lstlisting}

The front-end puts the initialization clause in the block it's currently writing
to, but generates a new block for the condition, then generates (without knowledge
of the loop body) a block for the final loop clause (this block is needed to jump
to when compiling a \verb+continue+ statement). It inserts this block after
it has generated the body block.

This would therefore be a possible translation:

\begin{lstlisting}
	/* enter the loop */
L0:	jmp(L1);
	/* continue or break from the loop */
L1:	split(cond, L2, L4);
	/* go to the final clause */
L2:	jmp(L3);
	/* no final clause, jump to loop start */
L3:	jmp(L1);
L4:	...
\end{lstlisting}

It can be noticed quite easily that \verb+L3+ only has one predecessor (\verb+L2+), and its
predecessor only one successor. It can therefore be merged with \verb+L2+:

\begin{lstlisting}
	/* enter the loop */
L0:	jmp(L1);
	/* continue or break from the loop */
L1:	split(cond, L2, L3);
	/* go to the final clause */
	/* no final clause, jump to loop start */
L2:	jmp(L1);
L3:	...
\end{lstlisting}

This turns out to be quite an interesting case, however; it can also be noticed
that this example might be optimized further, so the \verb+split+ in \verb+L1+
jumps to itself immediately. This is because \verb+L2+ is empty besides the \verb+jmp+
instruction: the condition for empty loops to be inlined is therefore more relaxed.

In fact, all empty blocks (except \verb+L0+ and empty blocks that are their own predecessor) can be inlined:

\begin{lstlisting}
L0:	jmp(L1);
L1:	split(cond, L1, L2);
L2:	...
\end{lstlisting}

\begin{thebibliography}{9}
  \bibitem{llvm_ir}
  \emph{LLVM Language Reference Manual}. (2014) Consulted on 2014/11/11,
  \url{http://llvm.org/docs/LangRef.html}

  \bibitem{cfld_cross}
  \emph{Constant folding and cross compilation}. (s.d.) Consulted on 2014/11/11,
  \url{http://en.wikipedia.org/wiki/Constant_folding#Constant_folding_and_cross_compilation}
\end{thebibliography}

\newpage
\appendix
\section{Implementation Details for acc}
